Stochastic Approximation Monte Carlo with a Dynamic Update Factor

### 2019-07-17 papers/histogram/sad.tex
made all of Michael Perlins suggestions that seemed relavent.

\begin{abstract}
  We present a new Monte-Carlo algorithm based on the Stochastic
  Approximation Monte Carlo (SAMC) algorithm for directly calculating
  the density of states. The proposed method is Stochastic
  Approximation with a Dynamic update factor (SAD)
  which dynamically adjusts the update factor $\gamma$ during the course of
  the simulation. We test this method on the square-well fluid and
  compare the convergence time and average entropy error for several
  related
  Monte-Carlo methods. We find that both the SAD and $1/t$-Wang-Landau ($1/t$-WL)
  methods rapidly converge to the
  correct density of states without the need for the user to specify an
  arbitrary tunable parameter $t_0$ as in the case of SAMC.  SAD requires
  as input the temperature range of interst, in contrast to
  $1/t$-WL, which requires that the user identify the interesting range
  of energies.  Thus SAD is more convenient when the range of energies is
  not known in advance.
\end{abstract}

@@

\section{Introduction}
Over the past several decades a number of flat histogram Monte-Carlo
simulation algorithms have been developed which calculate the
thermodynamic properties of various systems for over a range of temperatures.  This
development began with the original histogram method, which used a
single canonical Monte Carlo simulation to predict properties for
nearby temperatures~\cite{ferrenberg1988new}.  For large systems this
approach is limited to a narrow temperature range because a single
canonical simulation explores only a small range of energies.  This
led to a variety of ``flat'' (or ``broad'') histogram
methods~\cite{penna1996broad, penna1998broad, swendsen1999transition,
  wang2001determining, wang2001efficient, trebst2004optimizing}, which
explore a wider range of energies.  Another benefit, in
contrast with low-temperature canonical Monte Carlo, is that these
approaches cannot be trapped in a local energy minimum.

Wang and Landau introduced the most widely used flat histogram
algorithm (WL) that uses an update factor
and a statistical histogram to compute the density of states of a given
system~\cite{wang2001determining, wang2001efficient}.  While the method
is incredibly powerful, it has a few disadvantages. Firstly, it
requires the user to select the range
of energies to be studied~\cite{landau2004new}. This adds an additional
hurdle to its application to systems for which the interesting range
of energies is not known \emph{a priori}.  The simulation
violates detailed balance albiet briefly as the size of the violation decreases
with time, which complicates convergence analysis.  In fact, the error
in a WL computation the error has been demonstrated to saturate at a
non-zero value~\cite{yan2003fast}, i.e. the method does not converge to
the true density of states~\cite{belardinelli2007wang,
belardinelli2007fast, belardinelli2008analysis,
belardinelli2014intrinsic, singh2012density}.
%
Belardinelli and Pereyra demonstrated that allowing the update factor
to decrease faster than $1/t$ leads to
nonconvergence~\cite{belardinelli2007wang}.
%% Around the same time, the convergence rate for WL was being
%% explored~\cite{zhou2005understanding,lee2006convergence,
%%   belardinelli2007wang}.
%
This leads to their $1/t$-WL algorithm which ensures that the error
continues to decrease asymptotically as
$1/\sqrt{t}$~\cite{belardinelli2008analysis}.

@@

In this work, we have developed an improved algorithm based on SAMC
that does not require an array of non-physical, user-defined inputs
and therefore should be easily applicable to any system. The method
does require the user to define a temperature range of ($T_\text{min}$
to $T_\infty$).  We call this method SAD (Stochastic Approximation
with a Dynamic update factor), and will discuss it in detail in the
methods section. We compare its convergence properties with three
existing flat histogram methods: WL, $1/t$-WL, and SAMC.

As a test case, we consider the square-well fluid i.e. a system of
particles whose interactions are governed by a square-well
potential~\cite{singh2003surface, barker2004perturbationSW}.  The
square-well potential is an ideal test-bed as it is a simple model for
a liquid, which includes both attractive and repulsive
interactions~\cite{barker1967-SW-perturbation, vega1992phase}.  The
potential $U(\textbf{r})$ for such a system is given by
\begin{equation}
 U(\textbf{r})=\begin{cases} \infty &
 \lvert\textbf{r}\rvert< \sigma\\-\epsilon &
 \sigma<\lvert\textbf{r}\rvert<\lambda\sigma\\0 &
 \lvert\textbf{r}\rvert > \lambda\sigma\end{cases}
\end{equation}
where $\sigma$ is the hard-sphere diameter of the particle, $\lambda$ is the
reduced range of the potential well, and $\epsilon$ is its depth.
This model has the further advantage that because the energy is
discrete, binning is not required.

%The organization of this paper is as follows: In Section~\ref{sec:histogram}, we
%describe in detail flat-histogram methods.
%Section~\ref{sec:sad} introduces the dynamic stochastic approximation method.
%In Section~\ref{sec:results}, we discuss the performance of the various methods
%applied to the model system.

In this work, we compare four flat histogram methods.  We outline the
general workings of each algorithm that we developed in detail while
summarizing algorithms that were developed in other works.  The
following methods are discussed and simulated for the square-well
fluid: Wang-Landau (WL), $1/t$-Wang-Landau ($1/t$-WL), Stochastic
Approximation Monte-Carlo (SAMC), and SAD.

@@

In this section we will introduce four closely related flat histogram
methods which each rely on a weight function $w(E)$.  In these
algorithms, the probability of accepting a move is given by
\begin{equation}
	\mathcal{P}(E_\text{old} \rightarrow E_\text{new})
	= \min\left[1,\frac{w(E_\text{old})}{w(E_\text{new})}\right]
\end{equation}
which biases the simulation in favor of energies with low weights.  A
set of weights that are proportional to the density of states of the
system $D(E)$ will result in an entirely flat histogram.  Thus flat
histogram is a criteria for convergence for these methods.  To avoid
overflow error, since the weights may vary over more than a few
hundred orders of magnitude, the natural logarithm of the weights are
stored.  Since in the microcanonical ensemble the entropy is defined
as $S(E) \equiv k_B\ln(D(E))$, the logarithm of the weights can be
thought of as an approximation of the entropy.

Each approach uses a random walk in energy space to
estimate the density of states.  The core of these approaches
is to continuously update the weights at each step of the simulation
\begin{equation}
	\ln{w_{t+1}(E)}=\ln{w_{t}(E)}
	+\gamma_t
\end{equation}
where $t$ is number of the current move, $\gamma(t)$ is a
move-dependent update factor, and $E$ is the current energy.  This
update causes the random walk to avoid energies that have been
frequently sampled, leading to a rapid exploration of energy space.
This approach, however, violates detailed balance, due to the
acceptance probabilities changing with each move.  The severity of
this violation decreases as we decrease $\gamma_t$.  The four methods
differ primarily in how they schedule the decreasing of $\gamma_t$.

@@

The Wang-Landau approach~\cite{wang2001efficient,wang2001determining,
  landau2014guide} begins with $\gamma=1$, and then decreases $\gamma$
in discrete stages.  We track the number of moves ending at each
energy in a histogram.  When that histogram is sufficiently flat,
$\gamma$ is decreased by a specified factor of $\frac12$.  The
flatness is defined by the ratio between the minimum value of the
histogram and its average value.  When this flatness reaches a
specified threshold (typically 0.8), the $\gamma$ value is decreased
and the histogram is reset to zero.  This approach requires that the
energy range of interest be known in advance, and difficulties can
occur with this flatness criteria due to the fact that some energies
in this energy range might never be
sampled~\cite{haber2014transition}.  The entire process is repeated
until $\gamma$ reaches a desired cutoff.

The Wang-Landau approach thus has three parameters that need be
specified: the factor by which to decrease $\gamma$ when flatness is
acheived, the flatness criterion, and the cutoff that determines when
the computation is complete.  In addition, an energy range (or in
general, a \emph{set} of energies) must be supplied, so that the
flatness criterion can be defined.

While this approach is very efficient and has been widely used, it
suffers a few shortcomings.  Firstly, the set of energies must be
specified ~\cite{wang2001efficient, schulz2003avoiding, yan2003fast},
which may require multiple simulations. Secondly, while Wang-Landau
converges quickly, it does not in general converge to the true density
of states ~\cite{belardinelli2008analysis, zhou2008optimal}.  It can
and does decrease $\gamma$ so quickly that it will never (for any
cutoff value) decrease the error in the density of states beyond a
given nonzero value.

\subsection{$1/t$-Wang-Landau}
The error saturation of Wang-Landau algorithm algorithm can be
corrected by modifying the update factor such that it does not
decrease too quickly.  Belardinelli and Pereyra raised the question as
to whether the update factor should be decrease by $\nicefrac12$ or
some other factor for each update~\cite{belardinelli2007fast}. They
implemented a schedule that enforced that if the update factor
$\gamma$ is ever less than $\nicefrac{1}{t}$ then the update factor
is set to $\nicefrac{1}{t}$ and the histogram is no longer tracked. Employing this
scheduler, they found that the error saturation is avoided since the
correct density of states is approached asymptotically as
$t^{-\frac12}$~\cite{belardinelli2008analysis}. Zhou et. al further
confirmed that the WL algorithm never converges exponentially and
succesfully bounded the statistical error between $t^{-\frac12}$ and
$1/t$~\cite{zhou2008optimal}.

Schneider et. al outline minor refinements to the $1/t$-WL algorithm
including $\nicefrac{N_S}{t}$ scaling and switching from standard WL
to $1/t$-WL when the update factor $\gamma <
\nicefrac{N_S}{t}$~\cite{schneider2017convergence}. As per the
original $1/t$ implementation~\cite{belardinelli2007fast}, the update
factor is decreased once all $N_S$ energy states are visited at least
once (i.e. $H(E) \neq 0$) effectively avoiding the concept of
`flatness'.  In this work, we refer to this refined algorithm as $1/t$-WL,
with the update factor asymptoting to $\nicefrac{N_S}{t}$.

\subsection{SAMC}
The Stochastic Approximation Monte Carlo (SAMC) algorithm addresses
the lack of convergence of Wang-Landau's approach with a simple
schedule by which the update factor $\gamma$ is continuously
decreased~\cite{liang2007stochastic, werlich2015stochastic,
  schneider2017convergence}.  The update factor is defined in the
original implementation~\cite{liang2007stochastic} in terms of an
arbitrary tunable parameter $t_0$,
\begin{align}
\gamma_{t}^{\text{SA}} =\frac{t_0}{\max(t_0,t)}
\end{align}
where as above $t$ is the number of moves that have been attempted.
SAMC offers extreme simplicity, combined with is proven convergence.
Provided the update factor satisfies
\begin{align}
\sum_{t=1}^\infty \gamma_{t} = \infty \quad\textrm{and}\quad
\sum_{t=1}^\infty \gamma_{t}^\zeta < \infty
\end{align}
where $\zeta \in \{1,2\}$, Liang has shown that the weights are proven
to converge to the true density of states~\cite{liang2006theory,
  liang2007stochastic}.  In addition, the energy range need not be
known a priori.  The time to converge depends only on the choice of
parameter $t_0$.  Unfortunately, there is no prescription for finding
an acceptable value for $t_0$, and while the algorithm formally
converges, for a poor choice of $t_0$ that convergence can be far too
slow to be practical. Liang \emph{et al.} give a rule of thumb in
which $t_0$ is chosen in the range from $2N_S$ to $100N_S$ where $N_S$
is the number of energy bins~\cite{liang2007stochastic}.  Schneider
\emph{et al.} found that for the Ising model this heuristic is helpful
for small spin systems, but that larger systems require an even higher
$t_0$ value~\cite{schneider2017convergence}.  We will describe below
one case we examined, in which $t_0$ needs to be as much as two orders
of magnitude higher than the rule of thumb of $100N_S$ in order to
converge in $10^{12}$ moves.

@@

We can place a rigorous \emph{lower} bound $t_{\min}$ on the number of
moves required to find the true entropy by considering the total change
that needs to be made to the entropy.
\begin{align}
  \Delta S_{\text{tot}} \equiv \sum_E S(E) - S_{\min}
\end{align}
The minimum number of moves that could converge the entropy is
the number of moves that will enable the above total entropy change,
which we can approximate using an integral:
\begin{align}
   \Delta S_{\text{tot}} &= \sum_{t=0}^{t_{\min}} \gamma_t \\
  % &= \int_0^{t_{\min}} \gamma(t)dt \\
  &= t_0 + \int_{t_0}^{t_{\min}} \frac{t_0}{t}dt
  \\
  &= t_0\left(1 + \ln\left(\frac{t_{\min}}{t_0}\right)\right)
\end{align}
Solving for ${t_{\min}}$ we find that
\begin{align}
  {t_{\min}} &= t_0 e^{\frac{\Delta S_{\text{tot}}}{t_0} - 1}
\end{align}
which means that the minimum time to convergence grows exponentially
as $t_0$ is made smaller.  You \emph{seriously} don't want to underestimate
$t_0$!

@@

\section{SAD Algorithm}\label{sec:sad}
The Stochastic Approximation with Dynamic update factor (SAD) method
is a variant of the SAMC Algorithm that attempts to dynamically choose
the modification factor rather than relying on system dependent
parameters such as $t_0$ or $\gamma_0$.  There is an immediate
advantage of such an algorithm where parameters are chosen independent
of system size or type. Each flat-histogram method has unique
advantages and disadvantages.  Wang-Landau and $1/t$-WL require an
energy range for initialization.  SAMC removes this energy range
requirement but requires simulating every possible energy. Our
proposed method SAD requires the user to input $T_\text{min}$, the
lowest temperature of interest, which is an immediate disadvantage of
the method. However, identifying a minimum temperature of interest
$T_\text{min}$ may be easier for a user than determining in advance
an energy rang of intereste or the unphysical parameter $t_0$.

@@

While for SAMC, the update factor is defined in the original
implementation, for SAD the update factor $\gamma_{t}^{\text{SAD}}$ is
thought of as $\nicefrac{\text{d}S}{\text{d}t}$. This tells us that
the SAMC parameter
$t_0$ should have dimensions of entropy.
We begin with an estimate of the average value of the entropy (relative
to the lowest entropy at $T_{\min}$).  If we assume a quadratic
dependence on energy (see Fig.~\ref{fig:entropy-cartoon}), this is given by
\begin{align}
\langle S\rangle \approx \frac13 \frac{E({T=\infty}) - E(T_{\min})}{T_{\min}}
\end{align}
We approximate this energy difference by $E_H -E_L$ where $E_H$ and
$E_L$ are defined below. The entropy numerator of the update factor in
general should scale with the total number of interesting energy states
$N_S$, since updates to the weights are distributed between that many
energy states.  The product $N_S\langle S\rangle$ is the total change
of entropy required (starting from constant weights) to find the true
entropy, and puts a lower bound on the convergence time. After
\emph{long} times, when all the energies have been a long time ago, we
wish for a lower update factor in order to more rapidly refine the
remaining error in entropy.  We track the time at which we first visted
each possible energy.  We define $t_L$ to be the last time that we
encountered an energy that we currently believe is in the energy range
of interest, so a $t\gg t_L$ we feel confident that we have established
the true energy range of interest. We gradually transition to a lower
update factor (but still asymptotically scaling as $\gamma_t \propto
1/t$ to ensure eventual convergence).  Finally, we wish for an update
factor that is \emph{never} greater than 1, because a very large update
factor could introduce very large errors in entropy that may take many iterations
to remove.  The SAD expression for $\gamma_t$ which incorporates these
ideas is:
\begin{align}
  \gamma_{t}^{\text{SAD}} =
     \frac{
       \frac{E_{H}-E_{L}}{T_{\text{min}}} + \frac{t}{t_L}
     }{
       \frac{E_{H}-E_{L}}{T_{\text{min}}} + \frac{t}{N_S}\frac{t}{t_L}
     }
\end{align}
where $E_H$ and $E_L$ are the current estimates for the highest and
lowest energies of interest as defined below.  This factor
asymptotically has the same $1/t$ behavior as the original SAMC
algorithm and with the same $N_S$ prefactor used by the $1/t$-WL
method; however for earlier values of $t$, the update factor drops as
$1/t^2$ and jumps every time a new energy is determined to be of
interest.  This behavior allows SAD to dynamically prevent the update
factor from decreasing too rapidly.

@@

Since SAD does not explore \emph{all} energy states, it needs to
determine what energy range corresponds to the temperature range of
interest defined by $T_{\min}<T<\infty$. The simulation is responsible
for determining and updating this energy range. Given the true entropy
$S(E)$, we can define the interesting energy range as
%\begin{align}
  $E(T_{\min}) <E< E(T=\infty)$
%\end{align}
where $E(T)$ is the energy that maximizes $S-E/T$.  During the course of the
simulation, this precise energy is challenging to evaluate accurately.
In order to ensure that we sample this entire energy range adequately,
we define two energy limits:  a high energy $E_H$ and a low
energy $E_L$, which define the range over which the energy histogram
is made flat. At move $t$, $E_H$ and $E_L$ are the greatest and lowest
energy that prior to that move
had the highest histogram value (i.e. been visited the most times) during
the course of the simulation.
This definition results in a ``ratcheting'' effect, in which $E_H$
may only increase, while $E_L$ may only decrease over the course of the
simulation, which results in a conservative estimate of the range of
energies that need be sampled.

During the simulation when considering a move inside of the energy
range of interest $E_L \le E \le E_H$, the weights are used as in the
three methods already described.  If $E\ge E_H$, the weight is taken
to be
\begin{align}\label{eq:highw}
  w(E>E_H) &= w(E_H),
\end{align}
which corresponds to an infinite temperature.  This choice ensures
that if the maximum in entropy is at an energy $E_{\max}>E_H$, then
the energy $E_{\max}$ will eventually have the highest number of
counts and the ratcheting will result in $E_H\ge E_{\max}$ .  At lower
energies, Boltzmann weights corresponding to the minimum temperature
are used:
\begin{align}\label{eq:loww}
  w(E<E_L) &= w(E_L)e^{-\frac{E_L-E}{T_{\min}}}.
\end{align}
This choice has the result that if the energy $E_{\min}$ at which the free
energy at $T_{\min}$ is minimized is less than $E_L$, the lower energy
limit will ratchet down to include $E_{\min}$.
Each time we change the value of $E_H$ or $E_L$, the weights within the
new portion of the interesting energy range are set to the expressions
in Equations~\ref{eq:highw} and~\ref{eq:loww}.

>>>>

\begin{abstract}
  We present a new Monte Carlo algorithm based on the Stochastic
  Approximation Monte Carlo (SAMC) algorithm for directly calculating
  the density of states. The proposed method is Stochastic
  Approximation with a Dynamic update factor (SAD)
  which dynamically adjusts the update factor $\gamma_t$ during the course of
  the simulation. We test this method on the square-well fluid and
  compare the convergence time and average entropy error for several
  related
  Monte Carlo methods. We find that both the SAD and $1/t$-Wang-Landau ($1/t$-WL)
  methods rapidly converge to the
  correct density of states without the need for the user to specify an
  arbitrary tunable parameter $t_0$ as in the case of SAMC.  SAD requires
  as input the temperature range of interst, in contrast to
  $1/t$-WL, which requires that the user identify the interesting range
  of energies.  Thus SAD is more convenient when the range of energies is
  not known in advance.
\end{abstract}

@@

\section{Introduction}
Over the past several decades a number of flat histogram Monte Carlo simulation
algorithms have been developed which calculate the thermodynamic properties of
various systems for over a range of temperatures.  This development began with
the original histogram method, which used a single canonical Monte Carlo
simulation to predict properties for nearby
temperatures~\cite{ferrenberg1988new}.  For large systems this approach is
limited to a narrow temperature range because a single canonical simulation
explores only a small range of energies.  This led to a variety of ``flat'' (or
``broad'') histogram methods~\cite{penna1996broad, penna1998broad,
swendsen1999transition, wang2001determining, wang2001efficient,
trebst2004optimizing}, which explore a wider range of energies.  Another
benefit, in contrast with low-temperature canonical Monte Carlo, is that these
approaches cannot be trapped in a local energy minimum.

Wang and Landau introduced the most widely used flat histogram algorithm (WL)
that uses an update factor and a statistical histogram to compute the density of
states of a given system~\cite{wang2001determining, wang2001efficient}.  While
the method is incredibly powerful, it has a few disadvantages. Firstly, it
requires the user to select the range of energies to be
studied~\cite{landau2004new}. This adds an additional hurdle to its application
to systems for which the interesting range of energies is not known \emph{a
priori}.  The simulation violates detailed balance, albeit briefly as the size of
the violation decreases with time, which complicates convergence analysis.  In
fact, the error in a WL computation has been demonstrated to saturate
at a non-zero value~\cite{yan2003fast}, i.e. the method does not converge to the
true density of states~\cite{belardinelli2007wang, belardinelli2007fast,
belardinelli2008analysis, belardinelli2014intrinsic, singh2012density}.
%
Belardinelli and Pereyra demonstrated that allowing the update factor to
decrease faster than $1/t$ leads to nonconvergence~\cite{belardinelli2007wang}
where $t$ corresponds to the number of moves.
%
This leads to their $1/t$-WL algorithm which ensures that the error continues to
decrease asymptotically as $1/\sqrt{t}$~\cite{belardinelli2008analysis}.

@@

In this work, we have developed an improved algorithm based on SAMC that does
not require an array of non-physical, user-defined inputs and therefore should
be easily applicable to any system. The method does require the user to define a
temperature range of interest ($T_\text{min}$ to $T_\infty$).  We call this
method SAD (Stochastic Approximation with a Dynamic update factor), and will
discuss it in detail in the methods section. We compare its convergence
properties with three existing flat histogram methods: WL, $1/t$-WL, and SAMC.

As a test case, we consider the square-well fluid i.e. a system of particles
whose interactions are governed by a square-well
potential~\cite{singh2003surface, barker2004perturbationSW}.  The square-well
potential is an ideal test-bed as it is a simple model for a liquid, which
includes both attractive and repulsive
interactions~\cite{barker1967-SW-perturbation, vega1992phase}.  The potential
$U(\textbf{r})$ for such a system is given by
\begin{equation}
 U(\textbf{r})=\begin{cases} \infty &
 \lvert\textbf{r}\rvert< \sigma\\-\epsilon &
 \sigma<\lvert\textbf{r}\rvert<\lambda\sigma\\0 &
 \lvert\textbf{r}\rvert > \lambda\sigma\end{cases}
\end{equation}
where $\sigma$ is the hard-sphere diameter of the particle, $\lambda$ is the
reduced range of the potential well, and $\epsilon$ is its depth. This model has
the further advantage that binning is not required because the energy is
discrete.

In this work, we compare four flat histogram methods.  We outline the
general workings of each algorithm that we developed in detail while
summarizing algorithms that were developed in other works.  The
following methods are discussed and simulated for the square-well
fluid: Wang-Landau (WL), $1/t$-Wang-Landau ($1/t$-WL), Stochastic
Approximation Monte Carlo (SAMC), and SAD.

@@

In this section we will introduce four closely related flat histogram
methods which each rely on a weight function $w(E)$.  In these
algorithms, the probability of accepting a move is given by
\begin{equation}
	\mathcal{P}(E_\text{old} \rightarrow E_\text{new})
	= \min\left[1,\frac{w(E_\text{old})}{w(E_\text{new})}\right]
\end{equation}
which biases the simulation in favor of energies with low weights.  A
set of weights that are proportional to the density of states of the
system $D(E)$ will result in an entirely flat histogram.  Thus flat
histogram is a criteria for convergence for these methods.  To avoid
overflow error, since the weights may vary over more than a few
hundred orders of magnitude, the natural logarithm of the weights are
stored.  Since in the microcanonical ensemble the entropy is defined
as $S(E) \equiv \ln(D(E))$ (where $k_B \equiv 1$), the logarithm of the weights can be
thought of as an approximation of the entropy.

Each approach uses a random walk in energy space to
estimate the density of states.  The core of these approaches
is to continuously update the weights at each step of the simulation
\begin{equation}
	\ln{w_{t+1}(E)}=\ln{w_{t}(E)}
	+\gamma_t
\end{equation}
where $t$ is number of the current move, $\gamma_t(t)$ is a
move-dependent update factor, and $E$ is the current energy.  This
update causes the random walk to avoid energies that have been
frequently sampled, leading to a rapid exploration of energy space.
This approach, however, violates detailed balance, due to the
acceptance probabilities changing with each move.  The severity of
this violation decreases as we decrease $\gamma_t$.  The four methods
differ primarily in how they schedule the decreasing of $\gamma_t$.

@@

The Wang-Landau approach~\cite{wang2001efficient,wang2001determining,
  landau2014guide} begins with $\gamma_t=1$, and then decreases $\gamma$
in discrete stages.  We track the number of moves ending at each
energy in a histogram.  When that histogram is sufficiently flat,
$\gamma_t$ is decreased by a specified factor of $\frac12$.  The
flatness is defined by the ratio between the minimum value of the
histogram and its average value.  When this flatness reaches a
specified threshold (typically 0.8), the $\gamma_t$ value is decreased
and the histogram is reset to zero.  This approach requires that the
energy range of interest be known in advance, and difficulties can
occur with this flatness criteria due to the fact that some energies
in this energy range might never be
sampled~\cite{haber2014transition}.  The entire process is repeated
until $\gamma_t$ reaches a desired cutoff.

The Wang-Landau approach thus has three parameters that need be
specified: the factor by which to decrease $\gamma_t$ when flatness is
acheived, the flatness criterion, and the cutoff that determines when
the computation is complete.  In addition, an energy range (or in
general, a \emph{set} of energies) must be supplied, so that the
flatness criterion can be defined.

While this approach is very efficient and has been widely used, it
suffers a few shortcomings.  Firstly, the set of energies must be
specified ~\cite{wang2001efficient, schulz2003avoiding, yan2003fast},
which may require multiple simulations. Secondly, while Wang-Landau
converges quickly, it does not in general converge to the true density
of states ~\cite{belardinelli2008analysis, zhou2008optimal}.  It can
and does decrease $\gamma_t$ so quickly that it will never (for any
cutoff value) decrease the error in the density of states beyond a
given nonzero value.

\subsection{$1/t$-Wang-Landau}
The error saturation of Wang-Landau algorithm algorithm can be
corrected by modifying the update factor such that it does not
decrease too quickly.  Belardinelli and Pereyra raised the question as
to whether the update factor should be decrease by $\nicefrac12$ or
some other factor for each update~\cite{belardinelli2007fast}. They
implemented a schedule that enforced that if the update factor
$\gamma_t$ is ever less than $\nicefrac{1}{t}$ then the update factor
is set to $\nicefrac{1}{t}$ and the histogram is no longer tracked. Employing this
scheduler, they found that the error saturation is avoided since the
correct density of states is approached asymptotically as
$t^{-\frac12}$~\cite{belardinelli2008analysis}. Zhou et. al further
confirmed that the WL algorithm never converges exponentially and
succesfully bounded the statistical error between $t^{-\frac12}$ and
$1/t$~\cite{zhou2008optimal}.

Schneider et. al outline minor refinements to the $1/t$-WL algorithm
including $\nicefrac{N_S}{t}$ scaling and switching from standard WL
to $1/t$-WL when the update factor $\gamma_t <
\nicefrac{N_S}{t}$~\cite{schneider2017convergence}. As per the
original $1/t$ implementation~\cite{belardinelli2007fast}, the update
factor is decreased once all $N_S$ energy states are visited at least
once (i.e. $H(E) \neq 0$) effectively avoiding the concept of
`flatness'.  In this work, we refer to this refined algorithm as $1/t$-WL,
with the update factor asymptoting to $\nicefrac{N_S}{t}$.

\subsection{SAMC}
The Stochastic Approximation Monte Carlo (SAMC) algorithm addresses
the lack of convergence of Wang-Landau's approach with a simple
schedule by which the update factor $\gamma_t$ is continuously
decreased~\cite{liang2007stochastic, werlich2015stochastic,
  schneider2017convergence}.  The update factor is defined in the
original implementation~\cite{liang2007stochastic} in terms of an
arbitrary tunable parameter $t_0$,
\begin{align}
\gamma_{t}^{\text{SA}} =\frac{t_0}{\max(t_0,t)}\label{eq:1}
\end{align}
where as above $t$ is the number of moves that have been attempted.
SAMC offers extreme simplicity, combined with is proven convergence.
Provided the update factor satisfies
\begin{align}
\sum_{t=1}^\infty \gamma_{t} = \infty \quad\textrm{and}\quad
\sum_{t=1}^\infty \gamma_{t}^\zeta < \infty
\end{align}
where $\zeta \in \{1,2\}$, Liang has shown that the weights are proven
to converge to the true density of states~\cite{liang2006theory,
  liang2007stochastic}.  In addition, the energy range need not be
known a priori.  The time to converge depends only on the choice of
parameter $t_0$.  Unfortunately, there is no prescription for finding
an acceptable value for $t_0$, and while the algorithm formally
converges, for a poor choice of $t_0$ that convergence can be far too
slow to be practical. Liang \emph{et al.} give a rule of thumb in
which $t_0$ is chosen in the range from $2N_S$ to $100N_S$ where $N_S$
is the number of energy bins~\cite{liang2007stochastic}.  Schneider
\emph{et al.} found that for the Ising model this heuristic is helpful
for small spin systems, but that larger systems require an even higher
$t_0$ value~\cite{schneider2017convergence}.  We will describe below
one case we examined, in which $t_0$ needs to be as much as two orders
of magnitude higher than the rule of thumb of $100N_S$ in order to
converge in $10^{12}$ moves.

@@

We can place a rigorous \emph{lower} bound $t_{\min}$ on the number of
moves required to find the true entropy by considering the total change
that needs to be made to the entropy.
\begin{align}
  \Delta S_{\text{tot}} \equiv \sum_E S(E) - S_{\min}
\end{align}
The minimum number of moves that could converge the entropy is
the number of moves that will enable the above total entropy change,
which we can approximate using an integral:
\begin{align}
   \Delta S_{\text{tot}} &= \sum_{t=0}^{t_{\min}} \gamma_t^{SA} \\
  % &= \int_0^{t_{\min}} \gamma(t)dt \\
  &= t_0 + \int_{t_0}^{t_{\min}} \frac{t_0}{t}dt
  \\
  &= t_0\left(1 + \ln\left(\frac{t_{\min}}{t_0}\right)\right)
\end{align}
Solving for ${t_{\min}}$ we find that
\begin{align}
  {t_{\min}} &= t_0 e^{\frac{\Delta S_{\text{tot}}}{t_0} - 1}
\end{align}
which means that the minimum time to convergence grows exponentially
as $t_0$ is made smaller.  You \emph{seriously} don't want to underestimate
$t_0$!

@@

\section{SAD Algorithm}\label{sec:sad}
The Stochastic Approximation with Dynamic update factor (SAD) method
is a variant of the SAMC Algorithm that attempts to dynamically choose
the modification factor rather than relying on system dependent
parameters such as $t_0$ or $\gamma_0$.  There is an immediate
advantage of such an algorithm where parameters are chosen independent
of system size or type. Each flat-histogram method has unique
advantages and disadvantages.  Wang-Landau and $1/t$-WL require an
energy range for initialization.  SAMC removes this energy range
requirement but requires simulating every possible energy. Our
proposed method SAD requires the user to input $T_\text{min}$, the
lowest temperature of interest, which is an immediate disadvantage of
the method. However, identifying a minimum temperature of interest
$T_\text{min}$ may be easier for a user than determining in advance
an energy range of interest or the unphysical parameter $t_0$.

@@

While for SAMC, the update factor is defined in the original
implementation (see equation~\eqref{eq:1}), for SAD the update factor $\gamma_{t}^{\text{SAD}}$ is
thought of as $\nicefrac{\text{d}S}{\text{d}t}$. This tells us that
the SAMC parameter
$t_0$ should have dimensions of entropy.
We begin with an estimate of the average value of the entropy (relative
to the lowest entropy at $T_{\min}$).  If we assume a quadratic
dependence on energy (see Fig.~\ref{fig:entropy-cartoon}), this is given by
\begin{align}
\langle S\rangle \approx \frac13 \frac{E({T=\infty}) - E(T_{\min})}{T_{\min}}
\end{align}
We approximate this energy difference by $E_H -E_L$ where $E_H$ and
$E_L$ are defined below. The entropy numerator of the update factor in
general should scale with the total number of interesting energy states
$N_S$, since updates to the weights are distributed between that many
energy states.  The product $N_S\langle S\rangle$ is the total change
of entropy required (starting from constant weights) to find the true
entropy, and puts a lower bound on the convergence time. After
\emph{long} times, when all the energies have been explored a long time ago, we
wish for a lower update factor in order to more rapidly refine the
remaining error in entropy.  We track the time at which we first visted
each possible energy.  We define $t_L$ to be the last time that we
encountered an energy that we currently believe is in the energy range
of interest, so a $t\gg t_L$ we feel confident that we have established
the true energy range of interest. We gradually transition to a lower
update factor (but still asymptotically scaling as $\gamma_t \propto
1/t$ to ensure eventual convergence).  Finally, we wish for an update
factor that is \emph{never} greater than 1, because a very large update
factor could introduce very large errors in entropy that may take many iterations
to remove.  The SAD expression for $\gamma_t$ which incorporates these
ideas is:
\begin{align}
  \gamma_{t}^{\text{SAD}} =
     \frac{
       \frac{E_{H}-E_{L}}{T_{\text{min}}} + \frac{t}{t_L}
     }{
       \frac{E_{H}-E_{L}}{T_{\text{min}}} + \frac{t}{N_S}\frac{t}{t_L}
     }
\end{align}
where $E_H$ and $E_L$ are the current estimates for the highest and
lowest energies of interest as defined below.  This factor
asymptotically has the same $1/t$ behavior as the original SAMC
algorithm and with the same $N_S$ prefactor used by the $1/t$-WL
method; however for earlier values of $t$, the update factor drops as
$1/t^2$ and jumps every time a new energy is determined to be of
interest.  This behavior allows SAD to dynamically prevent the update
factor from decreasing too rapidly.

@@

Since SAD does not explore \emph{all} energy states, it needs to
determine what energy range corresponds to the temperature range of
interest defined by $T_{\min}<T<\infty$. The simulation is responsible
for determining and updating this energy range. Given the true entropy
$S(E)$, we can define the interesting energy range as
%\begin{align}
  $E(T_{\min}) <E< E(T=\infty)$
%\end{align}
where $E(T)$ is the energy that maximizes $S-E/T$.  During the course of the
simulation, this precise energy is challenging to evaluate accurately.
In order to ensure that we sample this entire energy range adequately,
we define two energy limits:  a high energy $E_H$ and a low
energy $E_L$, which define the range over which the energy histogram
is made flat. At move $t$, $E_H$ and $E_L$ are the greatest and lowest
energy that prior to that move
had the highest histogram value (i.e. been visited the most times) at some point
during the course of the simulation.
This definition results in a ``ratcheting'' effect, in which $E_H$
may only increase, while $E_L$ may only decrease over the course of the
simulation, which results in a conservative estimate of the range of
energies that need be sampled.

During the simulation when considering a move inside of the energy
range of interest $E_L \le E \le E_H$, the weights are used as in the
three methods already described.  If $E\ge E_H$, the weight is taken
to be
\begin{align}\label{eq:highw}
  w(E>E_H) &= w(E_H),
\end{align}
which corresponds to an infinite temperature.  This choice ensures
that if the maximum in entropy is at an energy $E_{\max}>E_H$, then
the energy $E_{\max}$ will eventually have the highest number of
counts and the ratcheting will result in $E_H\ge E_{\max}$ .  At lower
energies, Boltzmann weights corresponding to the minimum temperature
are used:
\begin{align}\label{eq:loww}
  w(E<E_L) &= w(E_L)e^{-\frac{E_L-E}{T_{\min}}}.
\end{align}
This choice has the result that if the energy $E_{\min}$ at which the free
energy at $T_{\min}$ is minimized is less than $E_L$, the lower energy
limit will ratchet down to include $E_{\min}$.
Each time we change the value of $E_H$ or $E_L$, the weights outside the
new portion of the interesting energy range are set to the expressions
in Equations~\ref{eq:highw} and~\ref{eq:loww}.
